---
title: "Homework 1 - Stat 5302"
author: "Bailey Perry"
due date: "September 14, 2017"
output: pdf_document
---

## Heights Dataset Problem Work:
This problem involves the Height data introduced in Section 1.1 of the text.

```{r, message=FALSE, warning=FALSE}
#Number 1.1: Construct histograms of dheight and mheight.  
# Does the data appear to be normally distributed?
library(alr4)
data(Heights)
#View the data and use ?Heights to investigate the background of it too
#View(Heights)

# Construct the histograms as required
hist(Heights$mheight, main="Mother's Height Histogram", xlab="Height (inches)")
hist(Heights$dheight, main="Daughter's Height Histogram", xlab="Height (inches)")
```
Q: Does the data appear to be normally distributed?
Answer: In regards to the histogram of Mother's height, the data seems to be normally distributed because it centers around 62.5 and the range falls around 55-70. Then the data itself follows a bell curve shape that points to normality.

For the histogram of Daughter's height, the data looks slightly less normal than that of the Mother's plot. The center seems to skew more towards 63 or so, and the tails show some more variation than an average normal distribution. Although the data follows a semi-bell shape, it looks to be a bit skewed.


```{r, message=FALSE}
#Number 1.2: Give the standard error of the average height for mothers. 
# Give the standard error when height is measured in feet.
momse <- sqrt(var(Heights$mheight))/sqrt(nrow(Heights))
#the above is in inches and is following the formula: sd/sqrt(N)
momse

#Height measured in feet would need to be divided by 12
momseft <- momse/12
momseft
```


```{r, message=FALSE}
#Number 1.3: Give a 90% confidence interval for E(mheight) and provide a brief interpretation.

#df = n-1
n <- nrow(Heights)
df <- n-1
upper <- mean(Heights$mheight) + qt(0.95, df)*momse
#qnorm of 95 is used because we need to do 90+(alpha/2)
lower <- mean(Heights$mheight) - qt(0.95, df)*momse
upper; lower
```
Q: Provide a brief interpretation

Answer: The above values can be interpretated as the following -- The sample mean of mheight falls within the interval (62.348, 62.557) 90% of the time. That is the definition of confidence interval evaluation.

```{r, message=FALSE}
#Number 1.4: Construct the p-value for the hypothesis that 
# E(mheight) = 62.75 inches versus the alternative that it is not.
t.test(Heights$mheight, alternative="two.sided", mu=62.75)
#The corresponding p value is 0.000003161

#A t test is used here, since we did not assume normality AND 
# the t test is known to have more conservative estimates so it is 
# overall better for this case. We are testing if the mean is equal to 
# a hypothesized value or not.

#Online source: 
#https://www.princeton.edu/politics/about/file-repository/public/Handout8(2).pdf
#Student's t distribution - " many people use t-test even when 
# the distribution of the data is not known to be normal because 
# this test gives the results more conservative than the test based on 
# the normal approximation"
```
```{r, message=FALSE}
#Number 1.5: Assume that mheight is normally distributed with population mean 62.5 inches 
# and population variance 5.5 inches. Find the following:

#Utilize z = x-mu/sd and then pnorm(z)
#a.The probability that a randomly selected mother has height less than 68 inches.
pr1 <- pnorm(68, 62.5, sd=sqrt(5.5))
#The corresponding probability is 99.049%
pr1

#b. The probability that a randomly selected mother has height between 60 and 65 inches.
pr2 <- (pnorm(65,62.5,sd=sqrt(5.5))-pnorm(60,62,sd=sqrt(5.5)))
#The corresponding probability is 65.99%
pr2

#c. The number v such that Pr(E(mheight)-v < mheight < E(mheight)+v) = 0.95.
#this relates directly to confidence intervals and where the mean lies
v = qnorm(0.975)*sqrt(5.5)/(sqrt(nrow(Heights)))
v

#d. The 0.25 quantile of mheight.
quantile(Heights$mheight)
#The 0.25 quantile is at 60.8

#e. The median and first quartile of mheight
median(Heights$mheight)
#The median is 62.4 inches
quantile(Heights$mheight)
#or you can use summary function/either way works
summary(Heights$mheight)
#The first quartile is at 60.8 inches

#f. A 95% confidence interval for the height of a randomly selected mother.
ranheight <- sample(Heights$mheight, 1)
upper <- ranheight + qnorm(0.975)*momse
#qnorm of 97.5 is used because we need to do 95+(alpha/2)
lower <- ranheight - qnorm(0.975)*momse
upper; lower
```    


```{r, message=FALSE}
#Number 1.6: Test the hypothesis that E(mheight) = E(dheight) 
# -- (t test procedure). 
# Caution: mheight and dheight are correlated.
#paired is true and accounts for the correlation because they are related and
# inherently the mother's height affects the daughter's
t.test(Heights$mheight, Heights$dheight, paired = TRUE)

#A t test is again used, since we did not assume normality AND 
# the t test is known to have more conservative estimates so it is overall
# better for this case.
```


```{r, message=FALSE, warning=FALSE}
#Number 1.4 (Textbook): Old Faithful Problem
data(oldfaith)
#View(oldfaith)

library(ggplot2)
#utilize ggplot
ggplot ( oldfaith, aes(y=Interval, x= Duration) ) +
  geom_point() +
  geom_smooth( method=lm, se=FALSE) +
  geom_smooth( method=loess, se= FALSE, color="red") +
  xlab( "Duration")   +
  ylab( "Interval")   +
  ggtitle ( " The Interval based on the Duration")

#The blue line is linear and the red line is curved to the data using a loess
# smoothing line method
```

There seem to be two distinct groups in the data for predicting interval based on duration. The lower duration seems to predict smaller intervals to the next eruption. Likewise, a higher duration seems to predict larger intervals to the next eruption. This would make sense intuitively since a bigger/longer eruption would presumably take longer to "recover" from, as in build up to the next eruption.